Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-01-24 LICS_2025/e_graphs_with_bindings/sections/appendix.tex
wip

\section{$\catname{SLat}$}
\label{sec:appendix:1}

@@

>>>>

\subsection{$\catname{SLat}$}
\label{sec:appendix:1}

@@

\subsection{Pushout in $\catname{EHyp}(\Sigma)$}
\label{sec:appendix:pushout}

In this section we will describe the construction of the pushout in $\catname{EHyp}(\Sigma)$.
TODO

\subsection{Interpretation of $\Sigma$-terms as cospans of hypergraphs}
\label{sec:appendix:interpretation}
\begin{figure*}
  \[
  \adjustbox{width=\textwidth}{
  \tikzfig{./figures/interpretation}
  }
  \]
  \caption{Base cases for $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$}
  \label{fig:base_cases}
\end{figure*}

The base cases for interpretation function $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$ is given in the Figure~\ref{fig:base_cases}.
Then it is defined inductively by letting $[f \otimes g] \Coloneqq [f] \otimes [g]$ and $[f;g] \Coloneqq [f];[g]$ and by freeness induces the corresponding functor.

### 2025-01-24 LICS_2025/e_graphs_with_bindings/sections/soundness.tex

\begin{definition}
    We say that a $\Sigma$-term $f$ rewrites ($\leadsto$) to a $\Sigma$-term $g$ modulo SMC-laws via a rewrite rule $\langle l, r \rangle$ if they are representable as
    \[
    f = c_{1};(id_{k} \otimes l);c_{2} \qquad g = c_{2};(id_{k} \otimes r);c_{2}~.
    \]
    These representations are so-called SMC normal forms.
\end{definition}

This notion generalises to rewriting of closed $\Sigma$-terms by introducing contexts (in string diagrammatic notation).

>>>>

\begin{definition}
\label{def:rewrite}
    We say that a $\Sigma$-term $f$ rewrites ($\leadsto$) to a $\Sigma$-term $g$ modulo SMC-laws via a rewrite rule $\langle l, r \rangle$ if they are representable as
    \[
    f = c_{1};(id_{k} \otimes l);c_{2} \qquad g = c_{2};(id_{k} \otimes r);c_{2}~.
    \]
    These representations are so-called SMC normal forms.
\end{definition}

This notion generalises to rewriting of closed $\Sigma$-terms

\begin{definition}
    We say that a closed-$\Sigma$-term $f$ rewrites ($\leadsto$) to a closed-$\Sigma$-term $g$ modulo SMC-laws via a rewrite rule $\langle l, r \rangle$ if they are representable as either in Definition~\ref{def:rewrite}, or as 
    \[
    f = c_{1};(id_{k} \otimes d);c_{2} \qquad g = c_{2};(id_{k} \otimes e);c_{2}
    \]
    and $d \leadsto_{\langle l, r \rangle} e$
\end{definition}

\begin{lemma}

    Every closed $\Sigma^{+}$ term $f$ can be equivalently represented as
    \[
    f_{1} + \ldots + f_{n}
    \] where none of $f_{i}$ contain the join operator.
\end{lemma}
\begin{proof}
Such forms are given by orienting $\catname{SLat}$-equations such that $+$ gets propagated to the top, as, for example $f \otimes (g + h) = f \otimes g + g \otimes h$.
Most notably, we can also escape lambda abstractions by using distributivity law~\ref{law:distributivity}.
\end{proof}

This takes us to the following definition of rewriting for closed $\Sigma^{+}$-terms.
\begin{definition}
    We say that a closed-$\Sigma^{+}$-term $f$ rewrites ($\leadsto$) to a closed-$\Sigma^{+}$-term $g$ modulo SMC-laws via a rewrite rule $\langle l, r \rangle$ if they are representable as
    \[
    f = f_{1} + \ldots + f_{n} \qquad g = f_{1} + \ldots + f_{n}
    \]
    such that there is a permutation $\sigma$ such that there exist indices $i,j$ and $\sigma(f_{i}) \leadsto f_{j}$.
\end{definition}

To formulate the correspondence between rewriting systems we first define interpretation of closed $\Sigma^{+}$ terms in $\WellTypedMdaEcospans$.
It follows by first defining the interpretation of generators by extending the interpretation of plain $\Sigma$ terms as morphisms in $\MdaCospans$ as given in Appendix~\ref{sec:appendix:interpretation}.
The only missing cases are $[\textsf{ev}_{A,B}]$ and $[\Lambda_{A,B,C}]$.

\begin{figure}
    \begin{subfigure}{0.45\linewidth}
\[
\adjustbox{scale=0.5}{
    \tikzfig{./figures/ev_interpretation}
}
\]
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.45\linewidth}
        \[
        \adjustbox{scale=0.5}{
            \tikzfig{./figures/lambda_interpretation}
        }
        \]
    \end{subfigure}

\end{figure}

Then we can make $\WellTypedMdaEcospans$ into closed $\catname{SLat}$ category by defining a join of two cospans, $\Lambda$ of a cospan, and a designated $\textsf{ev}$ cospan and introducing DPOI rewrite rules to make these construction satisfy the laws of closed $\catname{SLat}$ symmetric monoidal category.
The join of two cospans is defined as in Figure~\ref{fig:f+g} and the quotienting is done through DPOI rewrite schema rules one for each axiom.
For example, the distributivity rule for $\otimes$ is given by the following schema rule
\[
\adjustbox{width=\linewidth}{
\tikzfig{./figures/semilattice_rule_1}~.
}
\]

We collect all such rewrite schemas into a set $\mathcal{S}$ and then a category $\WellTypedMdaEcospans / \mathcal{S}$ is a closed $\catname{SLat}$ SMC.

\begin{proposition}

For two closed $\Sigma^{+}$-terms $f$ and $g$ and a $\mathcal{E}$-equation $l = r$  $f \leadsto g$ if and only if $\llbracket f \rrbracket \Rrightarrow{} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.
\end{proposition}
\begin{proof}

\end{proof}

\begin{figure}
\[
        \adjustbox{scale=0.5}{
            \tikzfig{./figures/f_plus_g_new}
        }
\]
\caption{$+$ of two morphisms in $\WellTypedMdaEcospans$}
\label{fig:f+g}
\end{figure}

### 2025-01-24 LICS_2025/monoidal_e_graphs/sections/abstract.tex

% The technique of equipping graphs with an equivalence relation, called equality saturation, has recently proved both powerful and practical in program optimisation, particularly for satisfiability modulo theory solvers. 
The technique of equality saturation, which equips graphs with an equivalence relation, has proven effective for program optimisation.
We give a categorical semantics to these structures, called e-graphs, in terms of Cartesian categories enriched over a category of semilattices.
This approach generalises to monoidal categories, which opens the door to new applications of e-graph techniques, from algebraic to monoidal theories.
Finally, we present a sound and complete combinatorial representation of morphisms in such a category,  based on a generalisation of hypergraphs which we call e-hypergraphs.
They have the usual advantage that many of their structural equations are absorbed into a general notion of isomorphism. 
The advantage of this new principled approach to e-graphs is that for the first time it is possible to support double-pushout (DPO) rewriting for these structures which constitutes the main contribution of this paper.

>>>>

% The technique of equipping graphs with an equivalence relation, called equality saturation, has recently proved both powerful and practical in program optimisation, particularly for satisfiability modulo theory solvers. 
The technique of \emph{equality saturation}, which equips graphs with an equivalence relation, has proven effective for program optimisation.
We give a categorical semantics to these structures, called \emph{e-graphs}, in terms of Cartesian categories enriched over the category of semilattices.
This approach generalises to monoidal categories, which opens the door to new applications of e-graph techniques, from algebraic to monoidal theories.
Finally, we present a sound and complete combinatorial representation of morphisms in such a category,  based on a generalisation of hypergraphs which we call \emph{e-hypergraphs}.
They have the usual advantage that many of their structural equations are absorbed into a general notion of isomorphism.
This new principled approach to e-graphs enables double-pushout (DPO) rewriting for these structures, which constitutes the main contribution of this paper.

### 2025-01-24 LICS_2025/monoidal_e_graphs/sections/conclusion.tex

We have seen how e-graphs over algebraic theories can be expressed in terms of Cartesian categories enriched over the category of semilattices. 
In fact, e-graphs are an instance of a more general construction that can account for monoidal theories in terms of SMCs,  similarly enriched.  
Having described a translation of e-graphs into this categorical framework, a combinatorial representation of morphisms in the free category is also provided. 
As expected, this representation factors out the structural SMC equations, while remaining sensitive to the complexity-relevant equations induced by the enrichment.  
The design of such a representation and its corresponding rewrite theory proved to be delicate, and, together with its proof of completeness, constitute the main technical contributions of the paper.

>>>>

We have seen how e-graphs over algebraic theories can be expressed in terms of Cartesian categories enriched over the category of semilattices. 
In fact, e-graphs are an instance of a more general construction that can account for monoidal theories in terms of SMCs,  similarly enriched.  
Having described a translation of e-graphs into this categorical framework, we also provide a combinatorial representation of morphisms in the free category.
As expected, this representation factors out the structural SMC equations, while remaining sensitive to the complexity-relevant equations induced by the enrichment.  
We then showed our theory to be sound and complete with respect to our categorical semantics.

### 2025-01-24 LICS_2025/monoidal_e_graphs/sections/e-hypergraphs.tex

\begin{figure}
    \[
    \scalebox{0.5}{
        \tikzfig{../figures/combinatorial_semantics/extended_cospan_example}
    }    
    \]
    \captionsetup{belowskip=-3ex, skip=0pt}
    \caption{Extended cospan example}
    \label{fig:e-cospan-example}
\end{figure}

@@

Consider two extended cospans below 
\[
\scalebox{0.75}{
    \tikzfig{../figures/combinatorial_semantics/isomorphic_e_cospans}
}
\]
and consider a relation 
\[
R = \{x R y \text{ if } f_{int}(x) \consistency f_{int}(y) \text{ for } x,y \in n'\}
\]
and let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}_{j=1}^{k}$. 
We get an analogous partition for $m'$, $n''$, $m''$.
Two extended cospans are isomorphic if there exist isomorphisms $\alpha$, $\beta$ and $\gamma$ making the above diagram commute and such that $\alpha$ and $\gamma$ preserve order within each $p_i$.
\end{definition}
Intuitively, such construction suggests that the position of the image of $n$ within $n'$ relative to the strict internal interface does not matter; and that when considering individual components of hierarchical edges with their respective portions of strict internal input and output interfaces as plain cospans of hypergraphs (by only considering top-level vertices and edges of these components) they should be isomorphic cospans of hypergraphs in the sense from~\cite{bonchi_string_2022-1}.
An example of an extended cospan can be seen in Figure~\ref{fig:e-cospan-example}.
The partition of the interfaces is given with colours.
The labels suggest how $f_{int}, f_{ext}$ (respectively, $g_{int}, g_{ext}$) act on vertices, with the ordering of vertices in the interfaces given left to right.
Examples of isomorphic and non-isomorphic extended cospans can be found in Appendix~\ref{sec:appendix:iso}.

>>>>

\begin{figure}
    \[
    \scalebox{0.5}{
        \tikzfig{../figures/combinatorial_semantics/extended_cospan_example}
    }    
    \]
    \captionsetup{belowskip=-3ex, skip=0pt}
    \caption{Morphism of $\Ecospans$}
    \label{fig:e-cospan-example}
\end{figure}

@@

Consider two extended cospans below 
\[
\scalebox{0.75}{
    \tikzfig{../figures/combinatorial_semantics/isomorphic_e_cospans}
}
\]
and consider a relation 
\[
R = \{x R y \text{ if } f_{int}(x) \consistency f_{int}(y) \text{ for } x,y \in n'\}
\]
and let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}_{j=1}^{k}$. 
We get an analogous partition for $m'$, $n''$, $m''$.
Two extended cospans are isomorphic if there exist isomorphisms $\alpha$, $\beta$ and $\gamma$ making the above diagram commute and such that $\alpha$ and $\gamma$ preserve order within each $p_i$.
\end{definition}
An example of an extended cospan can be seen in Figure~\ref{fig:e-cospan-example}.
The partition of the interfaces is given with colours.
The labels suggest how $f_{int}, f_{ext}$ (respectively, $g_{int}, g_{ext}$) act on vertices, with the ordering of vertices in the interfaces given left to right.
Intuitively, such notion of isomorphism suggests that the position of the image of $n$ within $n'$ relative to the strict internal interface does not matter, for example, vertices
\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=red node, label={above:$v_{3}$}] (0) at (-0.5, 3.5) {};
		\node [style=red node, label={above:$v_4$}] (1) at (0, 3.5) {};
	\end{pgfonlayer}
\end{tikzpicture}
can be swapped with vertices
\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=blue node, label={above:$v_{5}$}] (0) at (-0.5, 3.5) {};
		\node [style=blue node, label={above:$v_6$}] (1) at (0, 3.5) {};
	\end{pgfonlayer}
\end{tikzpicture}
in the input interface yielding an isomorphic cospan.
However, we may not swap
\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=red node, label={above:$v_{3}$}] (0) at (-0.5, 3.5) {};
		\node [style=red node, label={above:$v_4$}] (1) at (0, 3.5) {};
	\end{pgfonlayer}
\end{tikzpicture}
with each other --- this would yield a non-isomorphic cospan.
Further examples of isomorphic and non-isomorphic extended cospans can be found in Appendix~\ref{sec:appendix:iso}.
